---
title: "Downstream analysis"
author: "Magali Eisik"
output:
  pdf_document:
    toc: true
    number_sections: true
---

```{r warning=FALSE,message=FALSE}
#Load require libraries
library(dplyr)      # Data manipulation
library(tidyr)      # Data cleaning / tidying
library(ggplot2)    # Data visualization
library(kableExtra) # Table formatting
library(knitr)      # Reporting tools
library(IlluminaHumanMethylation450kanno.ilmn12.hg19) # CpG annotation
library(clusterProfiler) # Pathway enrichment
library(org.Hs.eg.db)    # Gene IDs
library(enrichplot)      # Enrichment plots
library(readr)      # Read data
library(patchwork)  # Combine plots
library(kableExtra) # Table formatting
```

```{r}
#Read csv file with top20cpgs 
top_cpgs <- read.csv("Top20_CpGs_ElasticNet.csv", 
                     stringsAsFactors = FALSE)
#See first 3 rows
head(top_cpgs,3)
```

## Separate Cpgs by methylation status

```{r}
hyper_cpgs <- top_cpgs$CpG[top_cpgs$Methylation_Change == "Hypermethylated"]
hypo_cpgs <- top_cpgs$CpG[top_cpgs$Methylation_Change == "Hypomethylated"]

df_cpgs_met <- data.frame(
  CpG = c(hyper_cpgs, hypo_cpgs),
  Status = c(
    rep("Hypermethylated", length(hyper_cpgs)),
    rep("Hypomethylated", length(hypo_cpgs))
  )
)

df_cpgs_met %>%
  knitr::kable(
    format = "latex",
    escape = FALSE,
    booktabs = TRUE,
    caption = "CpG Sites by Methylation Status"
  ) %>%
  kableExtra::kable_styling(
    latex_options = c("striped", "hold_position")
  )
```

## CpG Annotation

```{r}
# The Illumina 450k annotation is loaded and converted 
#into a data frame to map each CpG site to its associated genomic information
ann450k <- getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)
ann450k_df <- as.data.frame(ann450k)
```

## Mapping CpG Sites to Their Associated Genes

```{r}
# Merge top CpGs with gene names
annotated_cpgs <- top_cpgs %>%
  left_join(ann450k_df[, c("Name", "UCSC_RefGene_Name")],
            by = c("CpG" = "Name"))
annotated_cpgs  
#UCSC_RefGene_Name contains the gene(s) associated with each CpG site 
#according to the UCSC RefGene annotation 


# Split multiple genes per CpG
df_annotation <- annotated_cpgs %>%
  tidyr::separate_rows(UCSC_RefGene_Name, sep = ";") %>%
  dplyr::rename(Gene = UCSC_RefGene_Name) %>%
  dplyr::filter(Gene != "") %>%
  dplyr::select(Methylation_Change, Gene)

```

## Gene Conversion for KEGG Analysis

```{r message=FALSE,warning=FALSE}

#separates genes by methylation status and converts 
#their symbols into ENTREZ IDs for downstream pathway enrichment analysis. 

# Hypermethylated 
hyper_genes <- unique(df_annotation$Gene[df_annotation$Methylation_Change == "Hypermethylated"])
hyper_ids <- bitr(hyper_genes, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)  #converts gene symbols to ENTREZ IDs using the human gene database
# OrgDb = org.Hs.eg.db--> homo sapiens data base
# Hypomethylated 
hypo_genes <- unique(df_annotation$Gene[df_annotation$Methylation_Change == "Hypomethylated"])
hypo_ids <- bitr(hypo_genes, fromType = "SYMBOL",
                 toType = "ENTREZID", OrgDb = org.Hs.eg.db)
```

## KEGG Enrichment Analysis by Methylation Status


KEGG pathway enrichment analysis was performed on the ENTREZ IDs of genes associated with hyper- and hypomethylated CpGs, identifying significantly overrepresented biological pathways with adjusted p-values, gene counts, and lists of contributing genes for each pathway

```{r}


# KEGG for hypermethylated
hyper_kegg <- enrichKEGG(gene = hyper_ids$ENTREZID, organism = 'hsa', pvalueCutoff = 0.05)

# KEGG for hypomethylated
hypo_kegg <- enrichKEGG(gene = hypo_ids$ENTREZID, organism = 'hsa', pvalueCutoff = 0.05)

#organism = 'hsa'--> homo sapiens
#pvalueCutoff = 0.05 →  only significantly enriched pathways whose p-value< 0.05 are considered.
```

## Data visualization

#Cnetplots (Category-Net Plot)

```{r warning=FALSE,message=FALSE}
cnetplot(hyper_kegg, showCategory = min(5, nrow(hyper_kegg)))
cnetplot(hypo_kegg, showCategory = min(5, nrow(hypo_kegg)))

```

The cnetplots reveal the relationship between genes and pathways, which are significantly enriched for hypermethylated and hypomethylated gene sets. According to the first cnetplot (hypermethylated genes), the most represented pathways include the cytoskeleton in muscle cells and type II diabetes mellitus. In the second plot, hypomethylated genes are concentrated in signaling and metabolic pathways, such as the phosphatidylinositol signaling system and inositol phosphate metabolism

## Dotplots

```{r}
prepare_top_kegg <- function(enrich_obj, top_n = 10) {
  enrich_obj@result %>%
    mutate(
      ID = as.character(ID),
      Description = as.character(Description),
      Count = as.numeric(Count),
      p.adjust = as.numeric(p.adjust)
    ) %>%
    arrange(p.adjust) %>%
    slice(1:top_n)
}


# Prepare top pathways

top_hyper <- prepare_top_kegg(hyper_kegg, top_n = 10)
top_hypo  <- prepare_top_kegg(hypo_kegg, top_n = 10)


# Create dotplot for hypermethylated CpGs

p_hyper <- ggplot(top_hyper, aes(x = reorder(Description, Count), y = Count)) +
  geom_point(aes(size = Count, color = -log10(p.adjust))) +
  coord_flip() +
  scale_color_gradient(low = "blue", high = "red") +
  scale_size_continuous(range = c(4, 10)) +  # Larger points
  labs(
    title = "KEGG Enriched Pathways (Hypermethylated CpGs)",
    x = "", 
    y = "Gene Count", 
    color = "-log10(adj p)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text.y = element_text(size = 10)
  )


# Create dotplot for hypomethylated CpGs

p_hypo <- ggplot(top_hypo, aes(x = reorder(Description, Count), y = Count)) +
  geom_point(aes(size = Count, color = -log10(p.adjust))) +
  coord_flip() +
  scale_color_gradient(low = "blue", high = "red") +
  scale_size_continuous(range = c(4, 10)) +
  labs(
    title = "KEGG Enriched Pathways (Hypomethylated CpGs)",
    x = "", 
    y = "Gene Count", 
    color = "-log10(adj p)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text.y = element_text(size = 10)
  )


ggsave("KEGG_Hypermethylated.png", plot = p_hyper, width = 10, height = 8, dpi = 300)
ggsave("KEGG_Hypomethylated.png", plot = p_hypo, width = 10, height = 8, dpi = 300)

ggplot(top_hyper, aes(x = reorder(Description, Count), y = Count)) +
  geom_point(aes(size = Count, color = -log10(p.adjust))) +
  coord_flip() +
  scale_color_gradient(low = "blue", high = "red") +
  scale_size_continuous(range = c(4, 10)) +  # Larger points
  labs(
    title = "KEGG Enriched Pathways (Hypermethylated CpGs)",
    x = "", 
    y = "Gene Count", 
    color = "-log10(adj p)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text.y = element_text(size = 10)
  )


ggplot(top_hypo, aes(x = reorder(Description, Count), y = Count)) +
  geom_point(aes(size = Count, color = -log10(p.adjust))) +
  coord_flip() +
  scale_color_gradient(low = "blue", high = "red") +
  scale_size_continuous(range = c(4, 10)) +
  labs(
    title = "KEGG Enriched Pathways (Hypomethylated CpGs)",
    x = "", 
    y = "Gene Count", 
    color = "-log10(adj p)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text.y = element_text(size = 10)
  )

```

The dot plots illustrate pathway-specific methylation changes. Hypermethylated CpGs are mainly enriched in pathways such as Cytoskeleton in muscle cells and Type II diabetes mellitus , whereas hypomethylated CpGs are enriched in Inositol phosphate metabolism and Phosphatidylinositol signaling system. Dot size corresponds to the number of genes in each pathway, color represents statistical significance (FDR), and gene ratio indicates the proportion of input genes associated with each pathway. These findings indicate that methylation changes selectively impact biological pathways, potentially repressing some while activating others.

## KEGG Pathway Enrichment and CpG Annotation by Methylation Status

```{r}
# Table with hyperlinks to KEGG pathways
# Hyper-methylated
hyper_table <- hyper_kegg@result
hyper_table$Methylation_Status <- "Hypermethylated"

# Hypo-methylated
hypo_table <- hypo_kegg@result
hypo_table$Methylation_Status <- "Hypomethylated"

# Combine tables
kegg_results <- rbind(
  hyper_table[, c("Description", "ID", "pvalue", "p.adjust", "Methylation_Status")],
  hypo_table[, c("Description", "ID", "pvalue", "p.adjust", "Methylation_Status")]
)

# Rename columns
colnames(kegg_results) <- c("Pathway_Name", "KEGG_ID", "PValue", "Adjusted_PValue", "Methylation_Status")


kegg_results <- kegg_results %>%
  mutate(KEGG_Link = paste0("https://www.kegg.jp/dbget-bin/www_bget?", KEGG_ID))


kegg_results <- kegg_results %>%
  mutate(Pathway_Name_LaTeX = paste0("\\href{", KEGG_Link, "}{", Pathway_Name, "}"))


kegg_results %>%
  dplyr::select(Pathway_Name_LaTeX, PValue, Adjusted_PValue, Methylation_Status) %>%
  knitr::kable(format = "latex", escape = FALSE, booktabs = TRUE,
               caption = "KEGG Pathways Enriched by Methylation Status",
               col.names = c("Pathway", "P-value", "Adjusted P-value", "Methylation Status"),
               align = "c") %>%
  kable_styling(latex_options = c("striped", "hold_position"), full_width = FALSE)
```

## Save results of KEGG analysis in CSV files

```{r}
# Hypermethylated
hyper_table <- hyper_kegg@result
hyper_table$Methylation_Status <- "Hypermethylated"

# Hypomethylated
hypo_table <- hypo_kegg@result
hypo_table$Methylation_Status <- "Hypomethylated"

# Combine tables
kegg_results <- rbind(
  hyper_table[, c("Description", "ID", "pvalue", "p.adjust", "Methylation_Status")],
  hypo_table[, c("Description", "ID", "pvalue", "p.adjust", "Methylation_Status")]
)

# Rename columns
colnames(kegg_results) <- c(
  "Pathway_Name", "KEGG_ID", "PValue", "Adjusted_PValue", "Methylation_Status")

# Create KEGG link column
kegg_results$KEGG_Link <- paste0("https://www.kegg.jp/dbget-bin/www_bget?", kegg_results$KEGG_ID)
kegg_results <- kegg_results[order(kegg_results$Adjusted_PValue), ]

# Save to CSV
write.csv(kegg_results, "KEGG_Pathways_Methylation.csv", row.names = FALSE)


#Gene names

final_cpg_genes <- annotated_cpgs %>%
  tidyr::separate_rows(UCSC_RefGene_Name, sep = ";") %>%
  dplyr::rename(Gene = UCSC_RefGene_Name) %>%
  dplyr::filter(Gene != "") %>%
  dplyr::select(CpG, Gene, Methylation_Change)


final_cpg_genes <- distinct(final_cpg_genes)

write.csv(final_cpg_genes,
          "Top20_CpGs_Annotated.csv",
          row.names = FALSE)
#Save csv

```

## Annotated CpGs with Associated Genes and Methylation Status

```{r}
#Display table with genes
cpgs_display <- final_cpg_genes %>% head(20)

cpgs_display %>%
  head(20) %>%
  knitr::kable(
    format = "latex",
    booktabs = TRUE,
    caption = "CpG Sites with Associated Genes and Methylation Status",
    col.names = c("CpG Site", "Associated Gene", "Methylation Status"),
    align = "c",
    escape = TRUE
  ) %>%
  kableExtra::kable_styling(
    latex_options = c("striped", "hold_position"),
    font_size = 11,
    full_width = FALSE
  )
cpgs_display %>%
  knitr::kable(format = "latex", escape = FALSE, booktabs = TRUE,
               caption = "Annotated Top 20 CpGs with Associated Genes and Methylation Status",
               col.names = c("CpG Site", "Associated Gene", "Methylation Status"),
               align = "c") %>%
  kable_styling(latex_options = c("striped", "hold_position"), font_size = 10)
```
