---
title: "Preparing GSE55763 Beta Value Datasets for Modeling (1000cpgs)"
author: "Magalí Eisik"
output:
  html_document:
    df_print: paged
---

This script processes the GSE55763 methylation dataset to create a tidy dataset with samples (GSM IDs) as rows, CpG probes as columns, and associated metadata including sex and age. The resulting dataset is ready for downstream modeling and analysis.

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

#Load require packages
library(tidyverse)
library(dplyr)
library(tibble)
library(tidyr)
library(readr)
```

Load metadata


```{r}
metadata_GSE55763 <- read.csv("GSE55763_metadata_clean.csv", header = TRUE, stringsAsFactors = FALSE)

# Display first 3 rows
head(metadata_GSE55763,3)
```



```{r, echo=FALSE}
metadata_GSE55763 <- metadata_GSE55763 %>%
  filter(grepl("population study", dataset))
dim(metadata_GSE55763)

```

This dataset contains the mapping between array IDs and their corresponding GSM sample IDs.

```{r}
array_to_GSM_mapping <- read.csv("array_to_GSM_mapping_clean.csv", header = TRUE, stringsAsFactors = FALSE)
head(array_to_GSM_mapping)

```




```{r,echo=FALSE}
# Keep only samples that are also present in the metadata
array_to_GSM_mapping <- array_to_GSM_mapping %>%
  filter(GSM_ID %in% metadata_GSE55763$X)

# Check dimensions again
dim(array_to_GSM_mapping)
```

The GSE55763_top1000_all_samples.csv dataset contains beta values for DNA methylation at the 1000 most variable CpG sites across all available samples. Unlike previous subsets, this dataset includes all 2663 samples, providing a comprehensive set for downstream analysis and modeling.

```{r}
betas_sub1000 <- read.csv("GSE55763_top1000_all_samples.csv", header = TRUE, stringsAsFactors = FALSE)
head(betas_sub1000 )
dim(betas_sub1000)

```

Reshapes betas_sub1000 from wide to long format, turning all columns except ID_REF into two columns: Array_ID (column names) and beta (values)



```{r,echo=FALSE}
# Remove 'X' prefix from column names
colnames(betas_sub1000) <- sub("^X", "", colnames(betas_sub1000))

# Keep only columns that exist in the mapping
array_cols <- intersect(colnames(betas_sub1000), array_to_GSM_mapping$Array_ID)

# Make sure there are columns left
length(array_cols)  # should be >0

# Filter dataset to keep only ID_REF + array columns
betas_sub1000_filtered <- betas_sub1000 %>%
  select(ID_REF, all_of(array_cols))
```




```{r}
betas_long <- betas_sub1000_filtered %>%
  pivot_longer(
    cols = -ID_REF,
    names_to = "Array_ID",
    values_to = "beta"
  ) %>%
  inner_join(array_to_GSM_mapping, by = "Array_ID")  # add GSM_ID

```

Map Array_ID to GSM_ID


```{r}
betas_wide <- betas_long %>%
  select(GSM_ID, ID_REF, beta) %>%
  pivot_wider(
    id_cols = GSM_ID,
    names_from = ID_REF,
    values_from = beta
  )

dim(betas_wide)  
head(betas_wide[,1:10])

```



The beta values were first reshaped from wide to long format, making each sample (GSM_ID) a row and each CpG probe a column, which prepares the data for analysis. The long-format beta data was then merged with the sample metadata using GSM_ID as the key, combining methylation measurements with relevant annotations. After verifying the dimensions and inspecting the first rows to ensure correctness, the final merged dataset was saved as a CSV file for downstream analyses or machine learning applications.

```{r}
betas_wide <- betas_long %>%
  select(-Array_ID) %>%
  pivot_wider(
    names_from = ID_REF,
    values_from = beta
  )
```

Merge with metadata

```{r}
metadata_filtered <- metadata_GSE55763 %>%
  filter(X %in% betas_wide$GSM_ID)

merged_data <- metadata_filtered %>%
  inner_join(betas_wide, by = c("X" = "GSM_ID"))
merged_data <- merged_data %>%
  select(-tissue, -dataset)

dim(merged_data)  
head(merged_data,3)
```

Save data set in CSV format

```{r}
write.csv(merged_data, "GSE55763_merged1000.csv", row.names = FALSE)