---
title: "KEGG Enrichment Analysis of Differentially Methylated CpG Sites"
author: "Magalí Eisik"
output:
  html_document:
    df_print: paged
---

After building the age prediction model and identifying the top 50 CpG sites that contribute most strongly to the prediction, we performed functional annotation using the KEGG pathway database. CpGs were classified as hypermethylated or hypomethylated based on their coefficients, and enrichment analysis was conducted to explore the biological pathways they are involved in.


```{r message=FALSE,warning=FALSE}
#Load require libraries
library(dplyr)
library(tidyr)
library(ggplot2)
library(kableExtra)
library(knitr)
library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)
library(readr)
library(kableExtra)
```

```{r}
#Read csv file with top50cpgs 
top_cpgs <- read.csv("C:/Users/Owner/Desktop/BINF7770/Top50_CpGs.csv", stringsAsFactors = FALSE)

#view first 5 rows
head(top_cpgs,50,5)
```

## Separate Cpgs: hyper- and hypomethylated 

```{r}
hyper_cpgs <- top_cpgs$CpG[top_cpgs$Methylation_Change == "Hypermethylated"]
hypo_cpgs  <- top_cpgs$CpG[top_cpgs$Methylation_Change == "Hypomethylated"]
```

## CpG Annotation

```{r}
# Load annotation
ann450k <- getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)
ann450k_df <- as.data.frame(ann450k)

# Merge top CpGs with gene names
annotated_cpgs <- top_cpgs %>%
  left_join(ann450k_df[, c("Name", "UCSC_RefGene_Name")],
            by = c("CpG" = "Name"))

# Split multiple genes per CpG
df_annotation <- annotated_cpgs %>%
  tidyr::separate_rows(UCSC_RefGene_Name, sep = ";") %>%
  dplyr::rename(Gene = UCSC_RefGene_Name) %>%
  dplyr::filter(Gene != "") %>%
  dplyr::select(Methylation_Change, Gene)
```



Gene Conversion for KEGG Analysis

```{r}
# Hypermethylated 
hyper_genes <- unique(df_annotation$Gene[df_annotation$Methylation_Change == "Hypermethylated"])
hyper_ids <- bitr(hyper_genes, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)

# Hypomethylated 
hypo_genes <- unique(df_annotation$Gene[df_annotation$Methylation_Change == "Hypomethylated"])
hypo_ids <- bitr(hypo_genes, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)
```

KEGG Enrichment Analysis by Methylation Status

```{r}
# KEGG for hypermethylated
hyper_kegg <- enrichKEGG(gene = hyper_ids$ENTREZID, organism = 'hsa', pvalueCutoff = 0.05)

# KEGG for hypomethylated
hypo_kegg <- enrichKEGG(gene = hypo_ids$ENTREZID, organism = 'hsa', pvalueCutoff = 0.05)
```


Data visualization

Cnetplots

```{r message=FALSE,warning=FALSE}

cnetplot(hyper_kegg, categorySize = "pvalue", showCategory = 5)
cnetplot(hypo_kegg, categorySize = "pvalue", showCategory = 5)

```


Dotplots 

```{r message=FALSE,warning=FALSE}

dotplot(hyper_kegg, showCategory = 10) + ggtitle("KEGG Enriched Pathways (Hypermethylated CpGs)")
dotplot(hypo_kegg, showCategory = 10) + ggtitle("KEGG Enriched Pathways (Hypomethylated CpGs)")

```

Barplots

```{r, message=FALSE, warning=FALSE}
barplot(hyper_kegg, showCategory = 10, title = "Hypermethylated CpG KEGG")
barplot(hypo_kegg, showCategory = 10, title = "Hypomethylated CpG KEGG")
```



KEGG Pathway Enrichment by Methylation Status

```{r}

# Hyper-methylated
hyper_table <- hyper_kegg@result
hyper_table$Methylation_Status <- "Hypermethylated"

# Hypo-methylated
hypo_table <- hypo_kegg@result
hypo_table$Methylation_Status <- "Hypomethylated"

# Combine tables
kegg_results <- rbind(
  hyper_table[, c("Description", "pvalue", "p.adjust", "Methylation_Status")],
  hypo_table[, c("Description", "pvalue", "p.adjust", "Methylation_Status")]
)

# Rename columns
colnames(kegg_results) <- c("Pathway_Name", "PValue", "Adjusted_PValue", "Methylation_Status")

# Add hyperlinks to KEGG pathways using the IDs (from the original object)
kegg_ids <- c(hyper_kegg@result$ID, hypo_kegg@result$ID)
kegg_results$Pathway_Name <- paste0(
  "<a href='https://www.kegg.jp/dbget-bin/www_bget?", kegg_ids, "' target='_blank'>",
  kegg_results$Pathway_Name,
  "</a>"
)


# Order by Adjusted_PValue
kegg_results <- kegg_results[order(kegg_results$Adjusted_PValue), ]

# Display table (paper-style)
kegg_results %>%
  knitr::kable("html", escape = FALSE,
               caption = "<b style='text-align:center;'>KEGG Pathways Enriched by Methylation Status</b>") %>%
  kableExtra::kable_styling(full_width = FALSE,
                            bootstrap_options = c("striped", "hover"),
                            position = "center")

```

Save results of KEGG analysis in CSV file

```{r}
# Hyper-methylated
hyper_table <- hyper_kegg@result
hyper_table$Methylation_Status <- "Hypermethylated"

# Hypo-methylated
hypo_table <- hypo_kegg@result
hypo_table$Methylation_Status <- "Hypomethylated"

# Combine tables
kegg_results <- rbind(
  hyper_table[, c("Description", "ID", "pvalue", "p.adjust", "Methylation_Status")],
  hypo_table[, c("Description", "ID", "pvalue", "p.adjust", "Methylation_Status")]
)

# Rename columns
colnames(kegg_results) <- c("Pathway_Name", "KEGG_ID", "PValue", "Adjusted_PValue", "Methylation_Status")

# Create KEGG link column
kegg_results$KEGG_Link <- paste0("https://www.kegg.jp/dbget-bin/www_bget?", kegg_results$KEGG_ID)

# Optional: remove "hsa" prefix from KEGG_ID if desired
kegg_results$KEGG_ID <- gsub("^hsa", "", kegg_results$KEGG_ID)

# Order by Adjusted_PValue
kegg_results <- kegg_results[order(kegg_results$Adjusted_PValue), ]

# Save to CSV
write.csv(kegg_results, "KEGG_Pathways_Methylation.csv", row.names = FALSE)

```

